---
title: "QRscore-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{QRscore-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# if (!requireNamespace("DESeq2", quietly = TRUE)) {
#   BiocManager::install("DESeq2")
# }
# if (!requireNamespace("BiocParallel", quietly = TRUE)) {
#   BiocManager::install("BiocParallel")
# }
library(QRscore)
library(DESeq2)
library(BiocParallel)
```


### Load the data and define age groups.
```{r}
data("example_dataset_raw_3000_genes")
```

```{r}

bulk_sparse_mat = example_dataset_raw_3000_genes$COUNTS
ages = example_dataset_raw_3000_genes$METADATA$AGE
ages[ages %in% c("20-29","30-39")] = "20-39"
ages[ages %in% c("40-49","50-59")] = "40-59"
ages[ages %in% c("60-69","70-79")] = "60-79"

```

### Data prefiltering and normalization

```{r message=FALSE, warning=FALSE}
col_means <- colMeans(bulk_sparse_mat, na.rm = TRUE)
col_zeros <- colMeans(bulk_sparse_mat==0, na.rm = TRUE)
col_ids <- which(col_means>5&col_zeros<0.2) # The threshold can be modified

bulk_df = bulk_sparse_mat[,col_ids]
bulk_df_inv = t(bulk_df)
coldata = data.frame(age = ages)
dds <- DESeqDataSetFromMatrix(countData = bulk_df_inv,
                                colData = coldata,
                                design = ~ age)
dds = estimateSizeFactors(dds)
normalized_mat <- counts(dds, normalized=TRUE)

```

```{r}
print("Number of kept genes after prefiltering:")
print(length(col_ids))
```

### Two-sample test (20-39 vs 60-79 age groups)

```{r}
kept_samples = coldata$age %in% c("20-39", "60-79")
normalized_mat_1 = normalized_mat[, kept_samples]
coldata_1 = coldata[kept_samples,]
results_2_sample <- QRscore.genetest(normalized_mat_1, coldata_1, pairwise_test = TRUE, pairwise_logFC = TRUE, test_mean = TRUE, test_dispersion = TRUE, num_cores = 2, approx = "asymptotic")
```

```{r}
gc()
```

#### Differentially expressed genes (DEGs) found by QRscore-Mean.

```{r}
results_2_sample_mean = results_2_sample$mean_test
results_2_sample_DEG = results_2_sample_mean[results_2_sample_mean$QRscore_Mean_adj_p_value<0.05,]
head(results_2_sample_DEG)
```

#### Differentially dispersed genes (DDGs) found by QRscore-Var.

```{r}
results_2_sample_var = results_2_sample$var_test
results_2_sample_DDG = results_2_sample_var[results_2_sample_var$QRscore_Var_adj_p_value<0.05,]
head(results_2_sample_DDG)
```


### Three-sample test

#### An example for only calculating the 3-sample test p-values for QRscore-Var

```{r}
results_3_sample <- QRscore.genetest(normalized_mat, coldata$age, pairwise_test = FALSE, pairwise_logFC = FALSE, test_mean = FALSE, test_dispersion = TRUE, num_cores = 2, approx = "asymptotic")
head(results_3_sample$var_test)                             
```

#### Full results: DDGs and DEGs and supplementary statistics, including pairwise 2-sample test p-values and fold changes.

```{r}
results_3_sample <- QRscore.genetest(normalized_mat, coldata$age, pairwise_test = TRUE, pairwise_logFC = TRUE, test_mean = TRUE, test_dispersion = TRUE, num_cores = 2, approx = "asymptotic")
head(results_3_sample$var_test) 
head(results_3_sample$mean_test) 
```

